---
title: "Tracking Pediatric LTFU + Aging Out | Kenya"
author: "Aaron Chafetz, Nelly Maina, and Lydia Odero"
format: docx
editor: source
---

## Background

Continuity on treatment (CoT) is vital for viral suppression of clients on treatment. CoT estimates among C/ALHIV are lower than adults. Aging-up may distort proxy CoT estimates. There is a need to take a deep-dive into the outcomes of CALHIV on treatment through individual level data to quantify the proportion accounted for by aging-up. The data used here come from the Kenya National Data Warehouse (NDW). Underlying code can be found on [GitHub](https://github.com/USAID-OHA-SI/Aging).

## Setup

We'll be working with two files from NDW's relational database - (1) a master client list and (2) a client visits log. We'll need to munge the master client list and merge as well as append that onto the visits log.

Before we import the data, we first will want to load the packages we'll be using. Please note that `tidyverse` is `v2.0.0`, which now includes the `lubridate` package (used in this analysis).

```{r}
#| label: dependencies 
# library(tidylog, warn.conflicts = FALSE) #explicitly referenced for detailing filtering/merging results
library(tidyverse)
library(scales, warn.conflicts = FALSE)
```

First we can import the master list...

```{r}
#| label: import_client
#| cache: true

clientlist <- read_csv("Data/AgingOut_Feb162023.csv",
                       col_types = list(.default = "c",
                                        `ART Initiation Date` = "D"))

clientlist <- clientlist %>% 
      rename(im=`Implementing Partner`, 
             facility=`Facility Name`, 
             mfl_code=`Facility MFL Code`,
             county=`County of Treatment`,
             id=`Client Number(De-identified)`, 
             sex=Sex, 
             orphan=Orphan, 
             residence=`County of Residence`,
             dob=DOB, 
             date_art_init=`ART Initiation Date`, 
             reg_art_init=`ART Regimen at Initiation`,
             line_art_init=`ART  Regimen Line at Initiation`, 
             reg_art_current=`Current ART Regimen`,
             line_art_current=`Current ART Regimen Line`)

glimpse(clientlist)

```

...and then the visit data.

```{r}
#| label: import_visits
#| cache: true

visits_data <- read_csv("Data/Appointments.csv", 
                        na = c("", "NA", "NULL"),
                        col_types = list(.default = "c"))

visits_data <- visits_data %>% 
  rename(im=`Implementing Partner`, 
         facility=`Facility Name`, 
         mfl_code=`Facility MFL Code`,
         id=`Client Number(De-identified)`, 
         appoint_date=`Appointment Date`, 
         visit_date=`Actual Visit Date`) 

glimpse(visits_data)
```

For simplicity, we will truncate number of characters in the id, making it easier to review, but still keeping it unique.

```{r}
#| label: ids

#capture the number of unique ids before truncating
  orig_unique <- length(unique(clientlist$id)) 

#truncate id to the last 9 characters in the client list
  clientlist <- clientlist %>% 
    mutate(id = str_sub(id, -9))

#are the truncated ids still unique?
  length(unique(clientlist$id)) == orig_unique
  
    paste("There are", length(unique(clientlist$id)) %>% comma, "unique clients")
    
#truncate id to the last 9 characters in the visit
  visits_data <- visits_data %>% 
    mutate(id =  str_sub(id, -9))
```

## Munging: Master Client List

Of the two datasets, we'll start by munging the master client list, `clientlist`. Looking at the data, we adjust the class of `DOB`, date of birth,  which is a string data-time variable and we want to store it as just as just a date.

```{r}
#| label: clients_dob

  #convert dob to date
  clientlist <- clientlist %>% 
    mutate(dob = as_date(dob) %>% ymd)
  
  glimpse(clientlist)
```

Within the master client list there are `r unique(clientlist$id) |> length() |> comma()` unique clients (`id`) despite there being `r nrow(clientlist) |> comma()`) records. The extra rows are a result of the client visiting multiple facilities over the course of their care. Since the meta client data may slightly vary (e.g. the birth date may be entered differently or the ART regimen line at initiation may not match), we need to just pick on of these rows of observation. We will use the visits dataset, `visits_data`, to identify the last visit for a given client and then use the information available at this facility to replace the rest of the client metadata.

```{r}
#| label: client_lastest

#identify the latest visit/current facility for a client
  curr_facility <- visits_data %>%
    mutate(visit_date = dmy(visit_date)) %>% 
    group_by(id) %>% 
    filter(visit_date == max(visit_date)) %>% 
    ungroup() %>% 
    distinct(id, mfl_code) %>% 
    mutate(current_facility = TRUE)

#ensure full list of ids after filtering
  length(unique(visits_data$id)) == length(unique(curr_facility$id))

#limit dataset to only key variables, remove other meta data info at earlier facilities and replace with current
clientlist <- clientlist %>%
  select(id, mfl_code, sex, dob, facility, county, date_art_init, reg_art_current) %>% 
  group_by(id) %>% 
  mutate(n = n()) %>% 
  ungroup() %>% 
  tidylog::left_join(curr_facility, by = c("id", "mfl_code")) %>% 
  mutate(current_facility = ifelse(is.na(current_facility), FALSE, current_facility),
         across(c(sex, dob, date_art_init, reg_art_current), \(x) case_when(current_facility | n == 1 ~ x))) %>% 
  group_by(id) %>% 
  fill(c(sex, dob, date_art_init, reg_art_current), .direction = "updown") %>% 
  ungroup() %>% 
  select(-c(n, current_facility))
```

```{r}
#| label: client_fac_dups
#| echo: false
multi_vals <- clientlist %>% 
  group_by(id, mfl_code) %>% 
  filter(n() > 1) %>% 
  ungroup() 

multi_clients <- multi_vals %>% 
  distinct(id) %>% 
  nrow()
```

If we review this client dataset, we notice that there are a `r nrow(multi_vals) ` cases in which the same `r multi_clients` client have  two or more different records within the same site with different meta data. To resolve this discrepancy, we will take the first observation.

```{r}
#| label: clients_remove

clientlist <- clientlist %>% 
  group_by(id, mfl_code) %>% 
  tidylog::filter(row_number() == 1) %>% 
  ungroup()
           
```

Another data issue we noticed is that `r filter(clientlist, dob > date_art_init) %>% nrow()` clients had ART initiation dates (`date_art_init`) before they were born (`dob`). We will exclude those clients from our master list.

```{r}
#| label: clients_wrong_init

clientlist <- clientlist %>% 
  tidylog::filter(dob <= date_art_init)

```

Lastly, for this analysis, we want to know when clients age out of the pediatric program (after they turn 15 years old). We need to create this variable based on the client's birthday (`dob`).

```{r}
#| label: clients_ageout

    master_clientlist <- clientlist %>% 
      mutate(date_age_out = dob %m+% years(15))

  glimpse(master_clientlist)
```
So at the end of the day, we have `r nrow(master_clientlist) %>%  comma()` unique client x facilitie pairs.

```{r}
#| label: clients_export
#| echo: false

write_csv(master_clientlist, "Dataout/master_clientlist.csv", na = "")
```

Before moving to the next step, we can quickly explore how many pediatric have been aging out (`date_age_out`) versus entering (`date_art_init`) each year for the past few years (excluding FY23 since we only have one quarter of data). This review is simply exploratory using the client master list data, and doesn't include LTFU.

```{r}
#| label: clients_explore
#| echo: false

master_clientlist %>% 
      mutate(date_art_init_fy = date_art_init %>% 
               quarter(with_year = TRUE, fiscal_start = 10) %>%
               str_replace("20", "FY") %>% 
               str_sub(end = 4),
             date_age_out_fy = date_age_out %>% 
               quarter(with_year = TRUE, fiscal_start = 10) %>%
               str_replace("20", "FY") %>% 
               str_sub(end = 4)) %>% 
      mutate(enter = date_art_init_fy,
             exit = date_age_out_fy) %>% 
      select(enter, exit) %>% 
      filter(!c(is.na(enter) & is.na(exit))) %>% 
      pivot_longer(everything(),
                   names_to = "type",
                   values_to = "fiscal_year",
                   values_drop_na = TRUE) %>% 
      count(type, fiscal_year, name = "value") %>% 
      pivot_wider(names_from = type) %>% 
      filter(fiscal_year %in% paste0("FY", 19:22))

```

## Munging: Visit Log

We have the relational master client list and now it's time to munge the visit log. We need to truncate the `id` like we did in the master client list so we can relate these two tables later on.

```{r}
#| label: visits_clean
    
  #clean up date
    visits_data <- visits_data %>% 
      mutate(across(c(visit_date, appoint_date), \(x) dmy(x)))
    
  #reorder columns for ease of review
    visits_data <- visits_data %>% 
      relocate(id, facility, .before = 1) %>% 
      relocate(visit_date, .before = appoint_date)
    
  glimpse(visits_data)
```

```{r}
#| label: visits_todrop
#| echo: false

ndrop <- visits_data %>% 
  anti_join(master_clientlist,
                     by = c("id", "facility", "mfl_code")) %>% 
  nrow() %>% 
  comma()
```

We want to restrict the dataset to only clients who appear in the master client list (and merge the client meta data onto the visit list), which means dropping `r ndrop` records that exist in the `visits_data` but not `master_clientlist`, i.e. clients have aged out.

```{r}
#| label: visits_semijoin
visits_data <- visits_data %>% 
  tidylog::semi_join(master_clientlist,
                     by = c("id", "facility", "mfl_code"))

```


```{r}
#| label: visits_date_missing
#| echo: false

df_appt_placeholder <- visits_data %>% 
  filter(appoint_date == "1900-01-01") 

df_appt_past <- visits_data %>% 
  filter(appoint_date <= visit_date)

v_appt_null <- visits_data %>% 
  filter(is.na(appoint_date)) %>% 
  nrow() %>% 
  comma()
```

If you review the data, you'll find some problems with appointment dates. One example is we can see appointments in there stored under `1900-01-01` possibly a way of coding missing data. There are `r nrow(df_appt_placeholder) |> comma()` records with this placeholder date (across `r unique(df_appt_placeholder$id) |> length() |> comma()` clients). We can also see a data entry problem with appointments if we look at the date compared with the visit date. The appointment dates should be in the future to identify the next visit date. There are, however, `r nrow(df_appt_past) |> comma()` observations for `r unique(df_appt_past$id) |> length() |> comma()` clients, where the appoint predates the visit. Additionally, `r v_appt_null` records have no appointment data. For those clients no correct or missing appoint dates, we have decided to drop these observations.

```{r}
#| label:  visits_problems
  #remove any visits that don't have a future appointment or an invalid date
 visits_data <-  visits_data %>%  
      tidylog::filter(!is.na(appoint_date), 
                      appoint_date > visit_date,
                      appoint_date != "1900-01-01")
```

There are also a number of duplicative records, which we will want to remove.

```{r}
#| label: visit_dups

clean_visits_data <- tidylog::distinct(visits_data) 

```

Before moving forward, we want to store the bounds of the visit date data in the dataset so we can filter out prior/future dates for LTFU and aging out.

```{r}
#| label:  visits_bounds
  min_visit <- min(clean_visits_data$visit_date, na.rm = TRUE)
  max_visit <- max(clean_visits_data$visit_date, na.rm = TRUE)
```
The visit observations occur between `r min_visit` and `r max_visit`.

## Join Master Client List to Visit Log

Now that we have clean data from both the client list and the vist log, we want to join these two relational tables (based on `id` and `facility`). Combining these data allow us to do two things further on: first to allow us to exclude any observations **after the client is 15 years old**; and second, to include the ART initiation date to capture "TX_NEW".

```{r}
#| label: join

  #merge on age data (for filtering)
    merged_visits_data <- clean_visits_data %>% 
      tidylog::left_join(master_clientlist,
                         by = c("id", "facility", "mfl_code"))
    
  #relocate patient data closer to id
    merged_visits_data <- merged_visits_data %>% 
      relocate(sex:date_age_out, .after = id)
    
    glimpse(merged_visits_data)

```

## Calculate Status

For the analysis, we need to have distinct rows for each date a status occurs. The next few sections will walk through the calculations for each status type.

### Status - Active

In order to calculate a client's status (active or lost to follow up, LTFU), we would ideally have the ART dispensing data to know how long before the client needs to return for more treatment. Without these data, we can create a proxy by using the length between visits, assuming their Rx is for this duration.

Let's start by including the next actual visit date onto the same line as the current visit.

```{r}
#| label: act_next

  #add in the next actual visit date for calculating duration between appointments
    merged_visits_data <- merged_visits_data %>% 
      arrange(id, visit_date) %>% 
      mutate(next_visit_date = lead(visit_date, order_by = visit_date),
             .by = id)

  glimpse(merged_visits_data)

```

Next, we can calculate the duration between actual visits. Based on PEPFAR's guidance, a patient would only be counted as lost 28 days after the planned gap (after missign theier scheduled appointment).

```{r}
#| label: act_duration

  #calculation duration between visits (proxy = 28 days after planned appt gap)
    merged_visits_data <- merged_visits_data %>% 
      mutate(visit_gap_planned = appoint_date - visit_date,
             visit_gap_allowed = appoint_date + days(28) - visit_date,
             visit_gap_actual = next_visit_date - visit_date) 

  glimpse(merged_visits_data)
```

Since this analysis is retrospective, in addition to knowing if a client is lost, we know whether the client has returned to treatment. We can calculate the following based on their visit gaps and if they have a future appointment.

**LTFU Logic**: Count a client as LTFU if (a) there is no future visit date (`next_visit-date`); or (b) the duration between visits exceeds (`visit_gap_actual`) the allowable visit gap (`visit_gap_allowed`); and (c) not if the client scheduled to return after we have actual visit data for in the dataset (`max_visit`)

**IIT/RTT Logic**: Count a client as interrupted and returned to treatment (RTT) if (a) they were originally LTFU; and (b) they have an actual next visit date (`next_visit_date`)

```{r}
#| label: act_status

  #ltfu (vs rtt) status
    merged_visits_data <- merged_visits_data %>% 
      mutate(ltfu = case_when(appoint_date + days(28) > max_visit ~ FALSE, #C 
                              is.na(next_visit_date) ~ TRUE, #A
                              visit_gap_actual > visit_gap_allowed ~ TRUE, #B 
                              TRUE ~ FALSE),
             rtt = case_when(ltfu == TRUE & is.na(next_visit_date) ~ FALSE, # A+B
                             ltfu & !is.na(next_visit_date) ~ TRUE))

  glimpse(merged_visits_data)

```

We do also want to remove any instances in the data where new initations are occurring after a patient has been to one or more visits. We will filter out any visits that occur before the `data_art_init`.

```{r}
#| label: act_init_lim

merged_visits_data <- merged_visits_data %>% 
  tidylog::filter(visit_date >= date_art_init)
```


At this stage, we will count all the patients as "Active" in the dataset. The patients will only become LTFU in the future after the `visit_gap_allowed` and will need to be added in as a separate row.

```{r}
#| label: act_df
  #create type for adding in different datasets
    merged_visits_data <- merged_visits_data %>% 
      mutate(status = "Active",
             date = visit_date)

  glimpse(merged_visits_data)
```

### Status - Aged Out

An important part of this analysis is tracking patients until they age out, i.e. are older than 15 years old. We will create a dataset (unique rows) to identify the date at which a client ages out. We'll filter down the master client list first to make sure we don't add age out dates after the observable period in the dataset and only if the client also exists in the visits log.

```{r}
#| label: ageout_df

  #add ageout date as a distinct record/row
    date_ageout <- master_clientlist %>% 
      filter(between(date_age_out, min_visit, max_visit)) %>% 
      filter(id %in% clean_visits_data$id) %>% 
      mutate(date = date_age_out,
             status = "Aged Out") 

glimpse(date_ageout)

```


### Status - LTFU

In the earlier section, we calculated when patients were active versus LTFU/ITT, but we need to actually append on distinct observations of when that occurs.

```{r}
#| label: ltfu_df
 
  #add in LTFU date as row (will be future period ahead of last visit)
    date_ltfu <- merged_visits_data %>%
      filter(ltfu == TRUE) %>%
      mutate(date = visit_date + visit_gap_allowed + days(1),
             status = ifelse(rtt == FALSE, "LTFU", "IIT")) %>%
      select(-visit_date:-visit_gap_actual) %>% 
      filter(between(date, min_visit, max_visit),
             date < date_age_out)

  glimpse(date_ltfu)
```

### Status - New

From the master client list, we can use the ART initiation date (`date_art_init`) to approximate TX_NEW as per PEPFAR MER guidance.

It's important to note a few things: (a) The initiations are separate from a separate source than the visit data; (b) if a client is initiated, they should also have a active status (i.e. have visits data), though this isn't always true in the data; (c) active status **should** be inclusive of new initiations (similar to MER where TX_CURR includes TX_NEW); and (d) some patients in data had initiation dates after one or many visits and these were removed in an earlier step.

```{r}
#| label: new_df

#include new initiation (between bounds of visit dataset and excluded 15+)
    status_new <- master_clientlist %>% 
      mutate(period = date_art_init %>% 
               quarter(with_year = TRUE, fiscal_start = 10) %>%
               str_replace("20", "FY") %>% 
               str_replace("\\.", "Q"),
             fiscal_year = period %>% 
               str_sub(3,4) %>% 
               paste0(20, .) %>% 
               as.integer(),
             status = "New",
             date = date_art_init) %>%
      tidylog::filter(date_age_out > date_art_init) %>% 
      tidylog::filter(id %in% clean_visits_data$id) %>% 
      tidylog::filter(between(date_art_init, min_visit, max_visit))

  glimpse(status_new)

```

## Appended Status Data

We have now have three separate data frames by status that need to be combined back together: Active, Aged Out, and LTFU/IIT. Once we bind these together, we need to remove any observations for clients who have aged out (+15). We also need to remove any patients who didn't have any data in the visits log, but were created by binding on the age out date.

```{r}
#| label: appended_df

  #bind date of ltfu date + age out date back onto dataset
    appended_visits_data <- merged_visits_data %>% 
      bind_rows(date_ltfu, date_ageout) %>% 
      arrange(id, date)

  #remove patients after 15+ years old
    appended_visits_data <- appended_visits_data %>% 
      tidylog::filter(date <= date_age_out)
  
  #remove patients where only observation is aging out
    appended_visits_data <- appended_visits_data %>% 
      group_by(id) %>% 
      tidylog::filter(!(status == "Aged Out" & n() == 1)) %>% 
      ungroup()
    
    glimpse(appended_visits_data)
```

### Quarter Status

At this point, we have combined all the status data together expect for new initiations (which are additive). We want to convert these calendar dates into status by fiscal quarter to approximate PEPFAR data, so we need to take the status of last observation for each fiscal quarter or year. First up, we need to apply the periods and fiscal year based on the status date and then arrange the status types so that Active comes after LTFU (in the event they both occur on the same day).

```{r}
#| label: status_cleaning

#apply fiscal quarter
    appended_visits_data <- appended_visits_data %>% 
      mutate(period = date %>% 
               quarter(with_year = TRUE, fiscal_start = 10) %>%
               str_replace("20", "FY") %>% 
               str_replace("\\.", "Q"),
             fiscal_year = period %>% 
               str_sub(3,4) %>% 
               paste0(20, .) %>% 
               as.integer())
    
  #arrange by status (Active to come after LTFU in event they occur on the same day)
    status_data <- appended_visits_data %>% 
      mutate(status = factor(status, c("LTFU", "IIT", "Active", "Aged Out"))) %>% 
      arrange(id, date, status)
    
    glimpse(status_data)
```

With the fiscal periods included, we can now

```{r}
#| label: status_qtr

  #filter by the last observed status per period to capture the status
    status_data_qtr <- status_data %>% 
      filter(row_number() == max(row_number()), 
             .by = c(id, period))

  #bind on new initations
    status_data_qtr <- status_data_qtr %>% 
      mutate(status = as.character(status)) %>% 
      bind_rows(status_new) %>%
      arrange(id, date)
    
    glimpse(status_data_qtr)
```

### Fiscal Year Status

We can repeat this as well to calculate each client's status at the end of the fiscal year.

```{r}
#| label: status_fy

    #filter by the last observed status per period to capture the status
    status_data_fy <- status_data %>% 
      filter(row_number() == max(row_number()),
             .by = c(id, fiscal_year))
    
    #bind on new initations
    status_data_fy <- status_data_fy %>% 
      mutate(status = as.character(status)) %>% 
      bind_rows(status_new) %>%
      arrange(id, date)
    
    glimpse(status_data_fy)
```

### Aggregate Tables

```{r}
#| label: status_tables

  #agg table of status by quarter (period)
  status_data_qtr %>% 
    count(fiscal_year, period, status) %>% 
    pivot_wider(names_from = status,
                values_from = n) %>% 
    relocate(New, Active, IIT, .after = 2)


  #agg table of status by fiscal_year 
    status_data_fy %>% 
      count(fiscal_year, status) %>% 
      pivot_wider(names_from = status,
                  values_from = n) %>% 
      relocate(New, Active, IIT, LTFU, .after = 2)
```

```{r}
#| label: visits_export
#| echo: false
#| include: false
#| eval: false

  #export
    write_csv(appended_visits_data, "Dataout/aging_patient-status_date.csv", na = "")
    zip("Dataout/aging_patient-status_date.zip",
        "Dataout/aging_patient-status_date.csv", extras = "-j")
    unlink("Dataout/aging_patient-status_date.csv")
    
    write_csv(status_data_qtr, "Dataout/aging_patient-status_qtr.csv", na = "")
    zip("Dataout/aging_patient-status_qtr.zip",
        "Dataout/aging_patient-status_qtr.csv", extras = "-j")
    unlink("Dataout/aging_patient-status_qtr.csv")
    
    write_csv(status_data_fy, "Dataout/aging_patient-status_fy.csv", na = "")
    zip("Dataout/aging_patient-status_fy.zip",
        "Dataout/aging_patient-status_fy.csv", extras = "-j")
    unlink("Dataout/aging_patient-status_fy.csv")
  

```

## Munging: Viz

One of the outputs of this analysis is to understand how clients are tracking each year. To best capture the ever changing nature of the client list, we can use a Sankey Diagram to visualize each year's changing enrollment data.

We'll need to start with the fiscal year status data (`status_data_fy`) and aggregate it if there are more than one row (this occurs due to appending on new initiations). We reorder New to the end so it is always last in the collapses text for our string parsing later on.

```{r}
#| label: viz_agg

  viz_collapsed <- status_data_fy %>% 
    mutate(status = fct_relevel(status, "New", after = Inf)) %>% 
    arrange(id, fiscal_year, status) %>% 
    mutate(status = as.character(status)) %>%
    summarise(status_start_yr = paste0(status, collapse = " - "),
              .by = c(id, fiscal_year))

  glimpse(viz_collapsed)
```

To make sure each patients exists each year and then pull the status forward of the next year.

```{r}
#| label: viz_next
  #expand each patient row to have every year and then pull the future year's status
  viz_collapsed <- viz_collapsed %>% 
    complete(fiscal_year, nesting(id)) %>% 
    mutate(status_end_yr = lead(status_start_yr, n = 1, order_by = fiscal_year),
           .by = id) 
  
  glimpse(viz_collapsed)
```

For each year, we want the start year to only start with TX_CURR as the base (and IIT and new initiations will exist mid way through), so we want to exclude the LTFU and Aged Out clients from the start year. Due to data entry problems, we need to exclude the new status in the end year where the patient already was enrolled in treatment.

```{r}
#| label: viz_years

  #removed LTFU from base year and get rid of "New" status from base year
  viz_collapsed <- viz_collapsed %>% 
    mutate(status_start_yr = 
             ifelse(str_detect(status_start_yr, "(LTFU|Aged Out)"), 
                    NA_character_, status_start_yr),
           status_start_yr = str_remove_all(status_start_yr, " - New")) %>% 
    filter(!(is.na(status_start_yr) & is.na(status_end_yr)))

 #move new to start for placement on plot (will move manually to mid point)
  viz_collapsed <- viz_collapsed %>% 
    mutate(status_start_yr = 
             ifelse(is.na(status_start_yr) & str_detect(status_end_yr, "New"),
                    "New", status_start_yr),
           status_end_yr = str_remove_all(status_end_yr, " - New"))
  
  glimpse(viz_collapsed)
```

Lastly, we want to aggregate the status categories by year and order them they way the should appear in the visualization (we will remove the numbers in post processing).

```{r}
#| label: viz_out

  #aggregate up by status groups for plot (and number for plot layout ordering)
  viz_out <- viz_collapsed %>% 
    count(fiscal_year, status_start_yr, status_end_yr, name = "patients", sort = TRUE) %>% 
    mutate(status_start_yr = case_match(status_start_yr,
                                        "Active" ~ "1. Active",
                                        "IIT" ~ "2. RTT",
                                        "New" ~ "3. New",
                                        # .default = status_start_yr), #check that no missing categories
                                        .default = "4. Unclassified"),
           status_end_yr = case_match(status_end_yr,
                                        "Active" ~ "1. Active",
                                        "IIT" ~ "2. IIT",
                                        "Aged Out" ~ "3. Aged Out",
                                        "LTFU" ~ "4. LTFU",
                                        # .default = status_end_yr) #check that no missing categories
                                        .default = "5. Unclassified")
    ) 

glimpse(viz_out)

```

Now are data are ready to be exported for us in aggregate level Sankey flow diagrams.

```{r}
#| label: viz_export
#| eval: false

  walk(2019:2022,
       ~ viz_out %>% 
         filter(fiscal_year == .x) %>% 
         write_csv(paste0("Dataout/aging-viz_fy", .x, ".csv"), na = "")
  ) 

```

From here, the data can be used in RawGraphs to create a Sankey plot. In RawGraphs, 1. load data (not saved by site) 2. Choose a chart: Alluvial Diagram 3. Mapping: Steps = status_start_year, status_end_year; Size = patients 4. Customize: width = 1280 height = 720 padding = 20 Links opacity = .8 5. Export: svg and cleaned up in Adobe Illustrator (this can also be done in PowerPoint)
